Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-12-16 ms/ms.Rmd
<edit index>

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

>>>>

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text (additional information included at https://seldenlab.github.io/elliptical.bifaces/). Elliptical bifaces have been found in all but one major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

### 2022-12-16 supp/index.qmd
<edit index>

@@

Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right. The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch. Landmarks and semilandmarks were applied in R [@R] using the `StereoMorph` library [@RN8973]. Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge, while others include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.

>>>>

Elliptical bifaces have a deep history in eastern North America extending from the Middle Archaic to well past European contact. In many instances they have been deposited in caches, which may represent large, corporate bundles decommissioned amid polity demise or radical transformations in social practice. Stylistic changes are evident in both length and width for these bifaces, and result in great variability in shape and size. While some undoubtedly were knapped for every-day, mundane uses, others were crafted from exotic cherts and reflect non-utilitarian uses. The _oversize_ or _exaggerated_ forms, mimicking combat weapons, are often referred to as symbolic weaponry. Such objects are typically those “whose size, delicacy, material of manufacture, and craftsmanship are outside the bounds of any weapon that might have been actually used” [@RN11177].

The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178]. In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken. Elliptical bifaces find their most dramatic expression as Caddoan and Mississippian exaggerated, symbolic weaponry. Mississippian Ramey knives, dating from c. CE 1000 to 1300 are found associated with Cahokia political dynamics. The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation. Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.

```{r figorget, fig.cap="Shell gorgets from sites in the American Southeast depicting elliptical bifaces."}

knitr::include_graphics('./images/composite.figure2.jpg')
```

Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm. Smaller elliptical biface forms, roughly half the size of the larger form, have also been identified at Spiro, along with several wooden implements that Early [-@RN1052] interpreted as memetic of elliptical bifaces. The bifaces from this latter category are the focus of this study, which is aimed at assessing whether resharpening and/or retouch may have resulted in convergent morphology at two sites in the ancestral Caddo area. A measure of morphological disparity is subsequently to assess whether biface morphology is more diverse at the larger size, which would suggest greater diversity in biface shape or size related to production activities.

@@

Prior to landmarking, elliptical bifaces from Millsap Cache and Jowell Farm were oriented with the most heavily retouched edge at top right. The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch. Landmarks and semilandmarks were applied in R 4.2.1 [@R] using the `StereoMorph` library v. 1.6.7 [@RN8973]. Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge, while others include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.
